Using Sorghum as an outgroup to find syntenic regions in Maize: regenerating

### 2017-10-20 SubGenome Analysis.Rmd
Interactive exploration with ggobi

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

# Import SynMap and GO Term Data
***
We parse the SynMap output file with ks/kn values (i.e. a DAGChainer file) to a flat format using a perl script (parseKsKnFile.pl) and import the data into R. Some ks and kn values are empty, which is to be expected. These are ignored for the purposes of finding the median, but included in the results if they are in a block that meets the cutoff criteria. GO terms related to Maize gene models are imported as well.

@@

# Choosing a KS cutoff value
***  
We need to select a cutoff value for ks (synonymous mutation rate) to differentiate between the most recent alpha duplication event and the previous beta duplication. Looking at ks value frequency (left), we should not see any obvious cutoff point.  By transforming ks values to block median values, we should see two (or more) peaks (center). In this (center) graph, the first peak represents orthologs and the second represents homeologs from "pregrass tetraploidy".  If the vertical line separates the first two peaks sufficently, it is probably a good enough value.  If not, we need to pick a new cutoff value for ks. Viewing the log10 transform of the ks values helps differentiate types of homologs (better version of this graph with color can be generated in CoGe).

@@

# Major Syntenic Blocks
***
Regions of synteny are not confined to a 1:1 relationship between Sorghum and Maize chromosomes. Where multiple regions in Maize are syntenic to the same region on Sorghum, we can assume this is a result of the whole genome duplication even.  These are the regions which will be sorted into subgenome 1 and subgenome 2.

@@

# SubGenomes
***
In order to separate subgenome 1 from subgenome 2, we implement a greedy algorithm to collect non-overlapping (when projected on sorghum) syntenic blocks by size. 
Psuedocode of the algorithm is as follows: foreach sorghum chromosome, set the largest syntenic chromosome (by gene) as sub1.  Foreach additional syntenic chromosome if doesn't overlap what is already in sub1, add it to sub1. Else, if it does not overlap anything in sub2, add to sub2.  Else toss.  After all syntenic chromosomes are process for this sorghum chromosome, count the number of genes in sub1 and sub2.  If sub2 is bigger, switch them.  *Toss out "small" syntenic chromosomes (51 genes or less?), and consider a tolerance for how much overlap must occur to consider it a true overlap.  Also consider switching to gene level overlap rather than whole-chromosome start/stop values.

@@

## Compared to Paper
How many chromosomes with synteny to sorghum were correctly placed in subgenome 1 or subgenome 2? Items in "Paper" only are missed subgenome assignments.  Items in "Greedy" only are false assignments.  Items in center are correct. (currently sorted by hand)

@@

# GO Terms in Maize
***

@@

# GO Term Comparison by Subgenome
***

@@

# Expression

@@

# Compare Homeolog Pairs

@@

# Expression and Abundance are Weakly Correlated

```{r subgenome_expression_protein_scatter, echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=8, fig.height=8, fig.align='center'}
## Start with the paired expression data
data <-
  homeologs.pairs %>%
  subset(Maize1 != "" & Maize2 != "") %>%
  select(Maize1, Maize2) %>%
  distinct() %>%
  inner_join(maize.expression.sample.avg, by=c("Maize1"="geneID")) %>%
  inner_join(maize.expression.sample.avg, by=c("Maize2"="geneID", "Sample"="Sample")) %>%
  subset(!is.na(FPKM_avg.x) & !is.na(FPKM_avg.y))
names(data)[4] <- "FPKM_maize1"
names(data)[5] <- "FPKM_maize2"

@@

## View as scatter
ggplot(data, aes(FPKM_maize1, dNSAF_maize1)) +
  geom_point(alpha=.1) +
  geom_smooth(method="loess", se=TRUE, fullrange=FALSE, level=0.95) +
  geom_abline(mapping = null, data = null, slope = 1, intercept = 0) +
  scale_x_log10() + scale_y_log10() +
  facet_wrap(~Sample, nrow=6) +
  labs(
    title = paste0("Scatterplot of Expression in Maize1 vs. Abundance in Maize1\nLog-Log Scale for gene pair(s).\nAlpha=.1, reference line (black), loess fit (blue)"),
    x = "Expression for Maize1 Gene (log scale)",
    y = "Abundance for Maize1 Gene (log scale)"
  ) 
```

```{r subgenome_expression_protein_foldchange_scatter, echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=6, fig.height=5, fig.align='center'}
## Start with the paired expression data
data <-
  homeologs.pairs %>%
  subset(Maize1 != "" & Maize2 != "") %>%
  select(Maize1, Maize2) %>%
  distinct() %>%
  inner_join(maize.expression.sample.avg, by=c("Maize1"="geneID")) %>%
  inner_join(maize.expression.sample.avg, by=c("Maize2"="geneID", "Sample"="Sample")) %>%
  subset(!is.na(FPKM_avg.x) & !is.na(FPKM_avg.y))
names(data)[4] <- "FPKM_maize1"
names(data)[5] <- "FPKM_maize2"

@@

data$foldChange_expr <- log2(data$FPKM_maize1) - log2(data$FPKM_maize2)
data$foldChange_abun <- log2(data$dNSAF_maize1) - log2(data$dNSAF_maize2)
ggplot(data, aes(foldChange_expr, foldChange_abun)) +
  geom_point(alpha=.1) +
  xlim(-10,10) + ylim(-10,10) +
  geom_smooth(method="lm", se=TRUE, fullrange=FALSE, level=0.95) +
  # facet_wrap(~Sample, nrow=6) +
  labs(
    title = paste0("Scatterplot of Foldchange in Expression in Maize1 vs.\nFoldchange in Abundance in Maize1\nAlpha=.1, linear fit, n=13413"),
    x = "log2(Expression Maize1) - log2(Expression Maize2)",
    y = "log2(Abundance Maize1) - log2(Abundance Maize2)"
  ) + 
  annotate("text",x=7,y=-8, label=paste("Pearson Cor = ", round(cor(data$foldChange_expr, data$foldChange_abun, method = c("pearson")),2)))
```

# Wrapup
***
Report generated on: `r as.character(format(Sys.Date(), format="%B %d, %Y"))`

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval.after = 'fig.cap')
```

@@

# 1 Import SynMap and GO Term Data
***
We parse the SynMap output file with ks/kn values (i.e. a DAGChainer file) to a flat format using a perl script (parseKsKnFile.pl) and import the data into R. Some ks and kn values are empty, which is to be expected. These are ignored for the purposes of finding the median, but included in the results if they are in a block that meets the cutoff criteria. GO terms related to Maize gene models are imported as well.

@@

# 2 Choosing a KS cutoff value
***  
We need to select a cutoff value for ks (synonymous mutation rate) to differentiate between the most recent alpha duplication event and the previous beta duplication. Looking at ks value frequency (left), we should not see any obvious cutoff point.  By transforming ks values to block median values, we should see two (or more) peaks (center). In this (center) graph, the first peak represents orthologs and the second represents homeologs from "pregrass tetraploidy".  If the vertical line separates the first two peaks sufficently, it is probably a good enough value.  If not, we need to pick a new cutoff value for ks. Viewing the log10 transform of the ks values helps differentiate types of homologs (better version of this graph with color can be generated in CoGe).

@@

# 3 Major Syntenic Blocks
***
Regions of synteny are not confined to a 1:1 relationship between Sorghum and Maize chromosomes. Where multiple regions in Maize are syntenic to the same region on Sorghum, we can assume this is a result of the whole genome duplication even.  These are the regions which will be sorted into subgenome 1 and subgenome 2.

@@

# 4 SubGenomes
***
In order to separate subgenome 1 from subgenome 2, we implement a greedy algorithm to collect non-overlapping (when projected on sorghum) syntenic blocks by size. 
Psuedocode of the algorithm is as follows: foreach sorghum chromosome, set the largest syntenic chromosome (by gene) as sub1.  Foreach additional syntenic chromosome if doesn't overlap what is already in sub1, add it to sub1. Else, if it does not overlap anything in sub2, add to sub2.  Else toss.  After all syntenic chromosomes are process for this sorghum chromosome, count the number of genes in sub1 and sub2.  If sub2 is bigger, switch them.  *Toss out "small" syntenic chromosomes (51 genes or less?), and consider a tolerance for how much overlap must occur to consider it a true overlap.  Also consider switching to gene level overlap rather than whole-chromosome start/stop values.

@@

## 5 Compared to Paper
How many chromosomes with synteny to sorghum were correctly placed in subgenome 1 or subgenome 2? Items in "Paper" only are missed subgenome assignments.  Items in "Greedy" only are false assignments.  Items in center are correct. (currently sorted by hand)

@@

# 6 GO Terms in Maize
***

@@

# 7 GO Term Comparison by Subgenome
***

@@

# 8 Expression

@@

# 9 Compare Homeolog Pairs

@@

# 10 Expression and Abundance are Weakly Correlated

```{r subgenome_expression_protein_scatter, echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=8, fig.height=8, fig.align='center', fig.cap = cap}
## Start with the paired expression data
data <-
  homeologs.pairs %>%
  subset(Maize1 != "" & Maize2 != "") %>%
  select(Maize1, Maize2) %>%
  distinct() %>%
  inner_join(maize.expression.sample.avg, by=c("Maize1"="geneID")) %>%
  inner_join(maize.expression.sample.avg, by=c("Maize2"="geneID", "Sample"="Sample")) %>%
  subset(!is.na(FPKM_avg.x) & !is.na(FPKM_avg.y))
names(data)[4] <- "FPKM_maize1"
names(data)[5] <- "FPKM_maize2"

@@

## View as scatter
ggplot(data, aes(FPKM_maize1, dNSAF_maize1)) +
  geom_point(alpha=.1) +
  geom_smooth(method="loess", se=TRUE, fullrange=FALSE, level=0.95) +
  geom_abline(mapping = null, data = null, slope = 1, intercept = 0) +
  scale_x_log10() + scale_y_log10() +
  facet_wrap(~Sample, nrow=6) +
  labs(
    title = paste0("Scatterplot of Expression in Maize1 vs. Abundance in Maize1\nLog-Log Scale for gene pair(s).\nAlpha=.1, reference line (black), loess fit (blue)"),
    x = "Expression for Maize1 Gene (log scale)",
    y = "Abundance for Maize1 Gene (log scale)"
  )

cap <- "Fig 10.1: "
```

```{r subgenome_expression_protein_foldchange_scatter, echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=6, fig.height=5, fig.align='center', fig.cap = cap}
## Start with the paired expression data
data <-
  homeologs.pairs %>%
  subset(Maize1 != "" & Maize2 != "") %>%
  select(Maize1, Maize2) %>%
  distinct() %>%
  inner_join(maize.expression.sample.avg, by=c("Maize1"="geneID")) %>%
  inner_join(maize.expression.sample.avg, by=c("Maize2"="geneID", "Sample"="Sample")) %>%
  subset(!is.na(FPKM_avg.x) & !is.na(FPKM_avg.y))
names(data)[4] <- "FPKM_maize1"
names(data)[5] <- "FPKM_maize2"

@@

data$foldChange_expr <- log2(data$FPKM_maize1) - log2(data$FPKM_maize2)
data$foldChange_abun <- log2(data$dNSAF_maize1) - log2(data$dNSAF_maize2)
ggplot(data, aes(foldChange_expr, foldChange_abun)) +
  geom_point(alpha=.1) +
  xlim(-10,10) + ylim(-10,10) +
  geom_smooth(method="lm", se=TRUE, fullrange=FALSE, level=0.95) +
  # facet_wrap(~Sample, nrow=6) +
  labs(
    title = paste0("Scatterplot of Foldchange in Expression in Maize1 vs.\nFoldchange in Abundance in Maize1\nAlpha=.1, linear fit, n=13413"),
    x = "log2(Expression Maize1) - log2(Expression Maize2)",
    y = "log2(Abundance Maize1) - log2(Abundance Maize2)"
  ) + 
  annotate("text",x=7,y=-8, label=paste("Pearson Cor = ", round(cor(data$foldChange_expr, data$foldChange_abun, method = c("pearson")),2)))

cap <- "Fig 10.2: Considering only homeologous pairs where both genes were measured for expression and protein abundance, we look at fold change from maize2 expression to maize1 expression vs. the fold change from maize2 abundance vs. maize2 abundance. A slightly positive trend indicates a mild positive correlation between an increase in expression and an increase in abundance."
```

# 11 Data Exploration

```{r echo=FALSE, out.width='100%'}
knitr::include_graphics('./Data/Images/High expressing genes tend to vary.png')
```

```{r echo=FALSE, out.width='100%'}
knitr::include_graphics('./Data/Images/Low Expressing Genes tend to stay low.png')
```

```{r echo=FALSE, out.width='100%'}
knitr::include_graphics('./Data/Images/Pollen is different.png')
```

```{r echo=FALSE, out.width='100%'}
knitr::include_graphics('./Data/Images/High-low variation.png')
```

Red Square = [Zm00001d020592](https://www.maizegdb.org/gene_center/gene/Zm00001d020592) Pericarp/Aleurone 27 DAP vs. [Zm00001d005793](https://www.maizegdb.org/gene_center/gene/Zm00001d005793)

Yellow Square = [Zm00001d003188](https://www.maizegdb.org/gene_center/gene/Zm00001d003188) Pericarp/Aleurone 27 DAP vs. [Zm00001d025753](https://www.maizegdb.org/gene_center/gene/Zm00001d025753)

Yellow X = Zm00001d039040 Mature Leaf 8

Pink Cross = Zm00001d015385 Mature Leaf 8

# 12 Wrapup
***
Report generated on: `r as.character(format(Sys.Date(), format="%B %d, %Y"))`
